Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-05 submission.tex
small changes to sec:data

%Increases in tagger accuracy does not readily translate to
%improved parsing.

@@

Similarly, \newcite{Mac:05} investigated the effects of PoS tag sets on tagger
performance in English, specifically the Wall Street Journal portion of the Penn
Treebank (PTB) \cite{Mar:San:Mar:93}. Based on linguistic
considerations \newcite{Mac:05} mapped the 45 tags of the original PTB tag set
 to more fine-grained tag sets to investigate whether additional linguistic information
could assist the tagger. Experimenting with both lexically
and syntactically conditioned modifications such as distinguishing between
count nouns and noncount nouns and between transitive and intransitive verbs,
they found that more fine-grained tag sets rarely led to improvements in tagger
accuracy; the most successful modification yielded an improvement in tagger
accuracy of 0.05 percentage points.  They did not find any statistically
significant improvements using linguistically informed distinctions, arguing
that their results do not support the hypothesis that it is possible to achieve
significant performance improvements in PoS tagging by utilizing a
finer-grained tag set.

% fixme:
Transition/overgang her ...

\section{The Norwegian Dependency Treebank}
\label{sec:data}
Our experiments our based on the newly developed Norwegian Dependency Treebank (NDT)
\cite{Sol:Skj:Ovr:14}, the first publicly available treebank for
Norwegian. It was developed at the National Library of Norway in
collaboration with the University of Oslo, and contains manually coded
syntactic and morphological annotation for
both varieties of Norwegian; 311 000 tokens of Bokmål and 303 000
tokens of Nynorsk. This paper only report results for Bokmål, the main variant.
The treebanked material mostly comprise newspaper
text, but also include government reports, parliament transcripts and
blog excerpts. The annotation process was supported by
the rule-based Oslo-Bergen Tagger \cite{Hag:Joh:Nok:00} and then manually
corrected by annotators, adding syntactic dependency analysis to the
morphosyntactic annotation.

\paragraph{Morphological Annotation}
The morphological annotation and PoS tag set of NDT is based on the same
inventory as used by the Oslo-Bergen
Tagger \cite{Hag:Joh:Nok:00,Sol:13}, which in turn is largely based on the work
of \newcite{Faa:Lie:Van:97}. The tag set consists of 12 morphosyntactic PoS
tags outlined in Table \ref{ndttagset}, with 7 additional tags for punctuation
and symbols. The tag set is thus rather coarse-grained, with broad categories
such as \texttt{subst} (noun) and \texttt{verb} (verb). The PoS tags are
complemented by a large set of morphological features, providing information
about morphological properties such as definiteness, number and tense.
Selected subsets of these
features are used in our tag set modifications, where the coarse PoS tag of
relevant tokens is concatenated with one or more features to include
more linguistic information in the tags.

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Tag} & \textbf{Description} \\
        \midrule
        \texttt{adj} & Adjective \\
        \texttt{adv} & Adverb \\
        \texttt{det} & Determiner\\
        \texttt{inf-merke} & Infinitive marker \\
        \texttt{interj} & Interjection \\
        \texttt{konj} & Conjunction \\
        \texttt{prep} & Preposition \\
        \texttt{pron} & Pronoun \\
        \texttt{sbu} & Subordinate conjunction \\
        \texttt{subst} & Noun \\
        \texttt{ukjent} & Unknown (foreign word) \\
        \texttt{verb} & Verb \\
        \bottomrule
    \end{tabular}
    \caption{Overview of the original PoS tag set of NDT (excluding punctuation tags).}
    \label{ndttagset}
\end{table}

\paragraph{Syntactic Annotation}
The syntactic annotation choices in NDT are largely based on the Norwegian
Reference Grammar \cite{Faa:Lie:Van:97}. Some central annotation choices are outlined in
Table \ref{ndtannotation}, taken from of \newcite{Sol:Skj:Ovr:14},
providing overview of the analyses of syntactic constructions that often
distinguish dependency treebanks, such as coordination and the treatment of
auxiliary and main verbs. The annotations comprise 29
dependency relations, including \textsc{adv} (adverbial), \textsc{subj}
(subject) and \textsc{koord} (coordination).

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Head} & \textbf{Dependent} \\
        \midrule
        Preposition & Prepositional complement \\
        Finite verb & Complementizer \\
        First conjunct & Subsequent conjuncts \\
        Finite auxiliary & Lexical/main verb \\
        Noun & Determiner \\
        \bottomrule
    \end{tabular}
    \caption{Central head-dependent annotation choices in NDT.}
    \label{ndtannotation}
\end{table}

@@

\paragraph{Data Set Split}
As there was no existing standardized data set split of NDT due to its recent
development, we first needed to define separate sections for training, development and
testing. Our proposed sectioning of the treebank follows a 
standard 80-10-10 split. In establishing the  
split, care has been taken to preserve contiguous texts in the various 
sections while also keeping them balanced in terms of genre.

Our defined split into sections for training, development testing, and held-out
testing will be distributed with the future releases of the treebank, but 
details for replicating the split will also be made available at a companion website
for the final version of the paper. 
% Our proposed split has already been used in the Norwegian contribution to the
% Universal Dependencies project \cite{Ovr:Hoh:16}.

\paragraph{Tagger}
As our experiments during development required many repeated cycles of training and
testing for the different modified tag sets, we sought a PoS tagger that is both
reasonably fast and accurate. There is a often a considerable trade-off between
the two factors, as the most accurate taggers tend to suffer in terms of speed
due to their complexity. However, a tagger that achieves both close to
state-of-the-art accuracy as well as very high speed is TnT \cite{Bra:00}. TnT
was furthermore recently used to evaluate the recently proposed universal tag
set \cite{Pet:Das:McD:12}. The sum of these factors led to TnT being the tagger
of choice for our experiments.

>>>>

@@

Similarly, \newcite{Mac:05} investigated the effects of PoS tag sets on tagger
performance in English, specifically the Wall Street Journal portion of the
Penn Treebank (PTB) \cite{Mar:San:Mar:93}. Based on linguistic considerations,
\newcite{Mac:05} mapped the 45 tags of the original PTB tag set to more
fine-grained tag sets to investigate whether additional linguistic information
could assist the tagger. Experimenting with both lexically and syntactically
conditioned modifications such as distinguishing between count nouns and
noncount nouns and between transitive and intransitive verbs, they found that
more fine-grained tag sets rarely led to improvements in tagger accuracy; the
most successful modification yielded an improvement in tagger accuracy of 0.05
percentage points.  They did not find any statistically significant
improvements using linguistically informed distinctions, arguing that their
results do not support the hypothesis that it is possible to achieve
significant performance improvements in PoS tagging by utilizing a
finer-grained tag set.

% fixme:
%Transition/overgang her ...

\section{The Norwegian Dependency Treebank}
\label{sec:data}
Our experiments are based on the newly developed Norwegian Dependency Treebank
(NDT) \cite{Sol:Skj:Ovr:14}, the first publicly available treebank for
Norwegian. It was developed at the National Library of Norway in collaboration
with the University of Oslo, and contains manually coded syntactic and
morphological annotation for both varieties of Norwegian; 311 000 tokens of
Bokmål and 303 000 tokens of Nynorsk. This paper only reports results for
Bokmål, the main variety. The treebanked material mostly comprise newspaper
text, but also include government reports, parliament transcripts and blog
excerpts. The annotation process was supported by the rule-based Oslo-Bergen
Tagger \cite{Hag:Joh:Nok:00} and then manually corrected by annotators, adding
syntactic dependency analysis to the morphosyntactic annotation.

\paragraph{Morphological Annotation}
\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Tag} & \textbf{Description} \\
        \midrule
        \texttt{adj} & Adjective \\
        \texttt{adv} & Adverb \\
        \texttt{det} & Determiner\\
        \texttt{inf-merke} & Infinitive marker \\
        \texttt{interj} & Interjection \\
        \texttt{konj} & Conjunction \\
        \texttt{prep} & Preposition \\
        \texttt{pron} & Pronoun \\
        \texttt{sbu} & Subordinate conjunction \\
        \texttt{subst} & Noun \\
        \texttt{ukjent} & Unknown (foreign word) \\
        \texttt{verb} & Verb \\
        \bottomrule
    \end{tabular}
    \caption{Overview of the original PoS tag set of NDT (excluding punctuation
        tags).}
    \label{ndttagset}
\end{table}

The morphological annotation and PoS tag set of NDT is based on the same
inventory as used by the Oslo-Bergen Tagger \cite{Hag:Joh:Nok:00,Sol:13}, which
in turn is largely based on the work of \newcite{Faa:Lie:Van:97}. The tag set
consists of 12 morphosyntactic PoS tags outlined in Table \ref{ndttagset}, with
7 additional tags for punctuation and symbols. The tag set is thus rather
coarse-grained, with broad categories such as \texttt{subst} (noun) and
\texttt{verb} (verb). The PoS tags are complemented by a large set of
morphological features, providing information about morphological properties
such as definiteness, number and tense.  Selected subsets of these features are
used in our tag set modifications, where the coarse PoS tag of relevant tokens
is concatenated with one or more features to include more linguistic
information in the tags.

\paragraph{Syntactic Annotation}
\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Head} & \textbf{Dependent} \\
        \midrule
        Preposition & Prepositional complement \\
        Finite verb & Complementizer \\
        First conjunct & Subsequent conjuncts \\
        Finite auxiliary & Lexical/main verb \\
        Noun & Determiner \\
        \bottomrule
    \end{tabular}
    \caption{Central head-dependent annotation choices in NDT.}
    \label{ndtannotation}
\end{table}

The syntactic annotation choices in NDT are largely based on the Norwegian
Reference Grammar \cite{Faa:Lie:Van:97}. Some central annotation choices are outlined in
Table \ref{ndtannotation}, taken from of \newcite{Sol:Skj:Ovr:14},
providing overview of the analyses of syntactic constructions that often
distinguish dependency treebanks, such as coordination and the treatment of
auxiliary and main verbs. The annotations comprise 29
dependency relations, including \textsc{adv} (adverbial), \textsc{subj}
(subject) and \textsc{koord} (coordination).

@@

\paragraph{Data Set Split}
As there was no standardized data set split of NDT due to its very recent
development, we needed to establish a data set split
(training/development/test). Our data set split of the treebank follows the
standard 80-10-10 (training/development/test) split and will be distributed
with the treebank and proposed as the new standard(?). In creating the data set
split, care has been taken to preserve contiguous texts in the various data
sets while keeping the split balanced in terms of genre. Our
proposed data set split was used in the Norwegian contribution to the Universal
Dependencies project \cite{Ovr:Hoh:16}. The split will be made available at a
companion website. In our experiments, we train on the training data and
evaluate on the development data, before finally evaluating our optimized tag
set on the held-out test data set.

\paragraph{Tagger}
For our experiments with tag set modifications, we sought a PoS tagger both
reasonably fast and accurate. There is a often a considerable trade-off between
the two factors, as the most accurate taggers tend to suffer in terms of speed
due to their complexity. However, a tagger that achieves both close to
state-of-the-art accuracy as well as very high speed is TnT \cite{Bra:00}. TnT
was furthermore recently used to evaluate the recently proposed universal tag
set \cite{Pet:Das:McD:12}. The sum of these factors led to TnT being the tagger
of choice for our experiments.
